---
title: "Chap3N"
author: "Kimaya Lacey"
format: docx
editor: visual
---

# Data Wrangling 

## installing packages 
```{r}
install.packages("tidyverse")
install.packages("dplyr")
install.packages("gapminder")

library(tidyverse)
library(dplyr)
library(gapminder)
```

## glimpse
```{r}
glimpse(gapminder)
head(gapminder)
dim(gapminder)
names(gapminder)
summary(gapminder)
```
## Prepare the example dateset 
```{r}
gapminder_data <- gapminder %>%
  rename(
    life_expectancy = lifeExp,
    population = pop,
    gdp_per_capita = gdpPercap
  )
```


## Selecting variables
```{r}
selected_data <- gapminder_data %>%
  select(
    country,
    continent,
    year,
    life_expectancy
  )

glimpse(selected_data)
```

## 
```{r}
reduced_data <- health_data %>%
  select(
    -participant_id,
    -free_text_response
  )
```

# Filtering Observations 

## filtering to one year 
```{r}
gapminder_2007 <- gapminder_data %>%
  filter(year == 2007)
```

## filter to one group 
```{r}
africa_2007 <- gapminder_data %>%
  filter(
    year == 2007, 
    continent == "Africa"
  )
```

## filtering using multiple categories
```{r}
selected_continents <- gapminder_data %>%
  filter(
    year == 2007,
    continent %in% c(
      "Africa",
      "Americas",
      "Europe"
    )
  )
```

# Creating new variables 

## create a population measure in millions
```{r}
gapminder_2007 <- gapminder_2007 %>%
  mutate(
    population_millions = population / 1e6
  )
```

## create categories
```{r}
gapminder_2007 <- gapminder_2007 %>%
  mutate(
    life_expectancy_group = case_when(
      life_expectancy < 60 ~ "Below 60 years",
      life_expectancy < 75 ~ "60 to 74 years",
      life_expectancy >= 75 ~ "75 years or more",
      TRUE ~ NA_character_
    )
  )
```

# Arranging observations 

## using arrange() function 
```{r}
highest_life_expectancy <- gapminder_2007 %>%
  arrange(
    desc(life_expectancy)
  ) %>%
  select(
    country,
    continent,
    life_expectancy
  )

head(
  highest_life_expectancy,
  n = 10
)
```

# Grouping and summarizing 

## mean life expectancy by continent
```{r}
continent_summary <- gapminder_2007 %>%
  group_by(continent) %>%
  summarize(
    mean_life_expectancy = mean(
      life_expectancy,
      na.rm = TRUE
    ),
    median_life_expectancy = median(
      life_expectancy,
      na.rm = TRUE
    ),
    country_count = n(),
    .groups = "drop"
  )

continent_summary
```

## visualize the summary 
```{r}
ggplot(
  data = continent_summary,
  mapping = aes(
    x = reorder(
      continent,
      mean_life_expectancy
    ),
    y = mean_life_expectancy
  )
) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Mean Life Expectancy by Continent",
    subtitle = "Country-level observations from 2007",
    x = NULL,
    y = "Mean life expectancy in years",
    caption = "Source: Gapminder"
  ) +
  theme_minimal()
```

# Wide and Long Data

## reshaping with pivot_longer
```{r}
vaccination_wide <- tibble(
  county = c(
    "County A",
    "County B",
    "County C"
  ),
  vaccination_2023 = c(
    62,
    71,
    58
  ),
  vaccination_2024 = c(
    68,
    73,
    64
  )
)

vaccination_wide
```

## pivot_longer
```{r}
vaccination_long <- vaccination_wide %>%
  pivot_longer(
    cols = starts_with("vaccination_"),
    names_to = "year",
    values_to = "vaccination_rate"
  ) %>%
  mutate(
    year = str_remove(
      year,
      "vaccination_"
    )
  )

vaccination_long
```

## visualize the long data
```{r}
ggplot(
  data = vaccination_long,
  mapping = aes(
    x = year,
    y = vaccination_rate,
    group = county,
    color = county
  )
) +
  geom_line(
    linewidth = 1
  ) +
  geom_point(
    size = 3
  ) +
  labs(
    title = "Vaccination Rates Increased Across Three Counties",
    x = "Year",
    y = "Vaccination rate (%)",
    color = "County"
  ) +
  theme_minimal()
```

## reshaping with pivot_wider 
```{r}
vaccination_wide_again <- vaccination_long %>%
  pivot_wider(
    names_from = year,
    values_from = vaccination_rate,
    names_prefix = "vaccination_"
  )

vaccination_wide_again
```

# Missing Data
```{r}
gapminder_data %>%
  summarize(
    missing_life_expectancy = sum(
      is.na(life_expectancy)
    ),
    missing_population = sum(
      is.na(population)
    ),
    missing_gdp = sum(
      is.na(gdp_per_capita)
    )
  )
```



