---
title: "Lab 2: Applying the Grammar of Graphics"
subtitle: "Student Version"
format:
  docx:
    toc: true
    toc-depth: 3
    code-copy: true
    code-tools: true
execute:
  eval: false
  warning: false
  message: false
---

# Learning Objectives

You will apply the Grammar of Graphics to several public health questions.

# Setup

Load `tidyverse` and `gapminder`.

```{r}
#| label: load-packages
library(tidyverse)
library(gapminder)
```

# Part 1: Prepare the Data

Create `gapminder_2007` by filtering to 2007 and renaming `lifeExp`, `gdpPercap`, and `pop` to descriptive names.
## 
```{r}
#| label: prepare-data
gapminder_data <- gapminder %>%
  rename(
    life_expectancy = lifeExp,
    population = pop,
    gdp_per_capita = gdpPercap
  )
gapminder_2007 <- gapminder_data %>%
  filter(year == 2007)
```

# Part 2: Scatterplot

Create a scatterplot of GDP per capita and life expectancy. Use a logarithmic x-axis, map continent to color, use an accessible palette, and include complete labels.

```{r}
#| label: scatterplot
#| fig-cap: "GDP per capita and life expectancy in 2007."
gapminder_07g <- gapminder_2007 %>% 
  ggplot(
    aes(
      x = gdp_per_capita, 
      y = life_expectancy, colour = continent
    )) + 
      geom_point() + 
      scale_x_log10() +
      geom_smooth(
        aes(group=1),
        method = "lm",
        se = FALSE
      ) +
      scale_color_viridis_d() +
      labs(
        title = "GDP per capita and life expectancy in 2007", 
        x = "Gross Domestic Product per Capita", 
        y = "Life Expectancy"
      )
gapminder_07g
```

# Part 3: Ranked Bar Chart

Calculate mean life expectancy by continent and create a ranked horizontal bar chart.

```{r}
#| label: bar-chart
cont_sum <- gapminder_2007 %>%
  group_by(continent) %>%
  summarize(
    mean_life_expectancy = mean(
      life_expectancy,
      na.rm = TRUE
    ),
    median_life_expectancy = median(
      life_expectancy,
      na.rm = TRUE
    ),
    country_count = n(),
    .groups = "drop"
  )
cont_sum
```

```{r}
#contsum_07c <- cont_sum
  
contsum_07c <- ggplot( 
  cont_sum,
    aes(
      x = mean_life_expectancy, 
      y = continent, fill = continent
    )
  ) + 
   geom_bar(
     stat = "summary",
     fun = "mean"
   ) +
  labs(
    title = "Mean Life Expectancy by Continent", 
    x = "Mean Life Expectancy",
    y = "Continent"
  )
contsum_07c
```

# Part 4: Distribution

Create a boxplot comparing life expectancy across continents.

```{r}
#| label: boxplot

# Add your code here.
```

# Part 5: Faceting

Create a faceted scatterplot showing the GDP–life expectancy relationship separately by continent.

```{r}
#| label: faceted-scatterplot

# Add your code here.
```

# Part 6: Identify the Grammar

For one graph, identify the data, aesthetics, geometry, scales, labels, facets, and theme.

**Response:**  

# Part 7: Independent Visualization

Create one additional chart that answers a public health question. State the question, justify the chart, use an informative title, and add alt text.

```{r}
#| label: independent-chart

# Add your code here.
```

# Reflection

1. Which aesthetic mappings changed the meaning of the graph?
2. When did faceting improve interpretation?
3. Which graph most clearly communicated one finding?

**Response:**  

# Deliverables

Submit the completed QMD, rendered HTML, five visualizations, reflections, and Git commit.
