---
title: "Lab 2: Applying the Grammar of Graphics"
subtitle: "Student Version"
format:
  docx:
    toc: true
    toc-depth: 3
    code-copy: true
    code-tools: true
execute:
  warning: false
  message: false
---

# Learning Objectives

You will apply the Grammar of Graphics to several public health questions.

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(gapminder)
```

# Part 1: Prepare the Data

```{r}
#| label: prepare-data
gapminder_data <- gapminder %>%
  rename(
    life_expectancy = lifeExp,
    population = pop,
    gdp_per_capita = gdpPercap
  )
gapminder_2007 <- gapminder_data %>%
  filter(year == 2007)
```

# Part 2: Scatterplot

```{r}
#| label: scatterplot
#| fig-cap: GDP per capita and life expectancy in 2007.
gapminder_07g <- gapminder_2007 %>% 
  ggplot(
    aes(
      x = gdp_per_capita, 
      y = life_expectancy, colour = continent
    )) + 
      geom_point() + 
      scale_x_log10() +
      geom_smooth(
        aes(group=1),
        method = "lm",
        se = FALSE
      ) +
      scale_color_viridis_d() +
      labs(
        title = "GDP per capita and life expectancy in 2007", 
        x = "Gross Domestic Product per Capita", 
        y = "Life Expectancy"
      )
gapminder_07g
```

# Part 3: Ranked Bar Chart

```{r}
#| label: bar-chart
cont_sum <- gapminder_2007 %>%
  group_by(continent) %>%
  summarize(
    mean_life_expectancy = mean(
      life_expectancy,
      na.rm = TRUE
    ),
    median_life_expectancy = median(
      life_expectancy,
      na.rm = TRUE
    ),
    country_count = n(),
    .groups = "drop"
  )
cont_sum
```

```{r}
#| label: horizontal bar chart
#contsum_07c <- cont_sum
  
contsum_07c <- ggplot( 
  cont_sum,
    aes(
      x = mean_life_expectancy, 
      y = continent, fill = continent
    )
  ) + 
   geom_bar(
     stat = "summary",
     fun = "mean"
   ) +
  labs(
    title = "Mean Life Expectancy by Continent", 
    x = "Mean Life Expectancy",
    y = "Continent"
  )
contsum_07c
```

# Part 4: Distribution

```{r}
#| label: boxplot
gpm_07bx <- ggplot(
  data = gapminder_2007,
  mapping = aes(
    x = continent,
    y = life_expectancy, colour = continent
  )
) +
  geom_boxplot() + 
  labs(
    title = "Life Expectancy by Continent", 
    x = "Continent",
    y = "Mean Life Expectancy"
    )

gpm_07bx
```

# Part 5: Faceting


```{r}
#| label: faceted-scatterplot
gpm_07fsc <- ggplot(
  data = gapminder_2007,
  mapping = aes(
    x = gdp_per_capita,
    y = life_expectancy, colour = continent
  )
) +
  geom_point(
    alpha = 0.7
  ) +
  scale_x_log10() +
  facet_wrap(
    vars(continent)
  ) + 
  labs(
    title = "Life Expectancy by Continent",
    subtitle = "Gapminder Data filtered to 2007",
    x = "Gross Domestic Product per capita",
    y = "Life Expectancy"
    ) + theme_light()
gpm_07fsc
```

# Part 6: Identify the Grammar

For one graph, identify the data, aesthetics, geometry, scales, labels, facets, and theme.

Faceted Scatterplot displaying the GDP per Capita and Life Expectancy Relationship. 
Data: Gapminder data from 2007

Aesthetics: x-axis: Gross Domestic Product (GDP) per capita
y-axis: Life expectancy

Geometry: Geom_point

Scales:
The x-axis uses a logarithmic scale for GDP per capita.
The y-axis shows life expectancy in years.
Color distinguishes the five continents.

Labels: 
Title: Life Expectancy by Continent
Subtitle: Gapminder Data
x-axis: Gross Domestic Product per capita
y-axis: Life Expectancy
Legend: continent

Facets: The graph is divided into five panels using the continent variable: Africa, Americas, Asia, Europe, and Oceania.
Theme: The graph uses a white background with gray grid lines.

# Part 7: Independent Visualization

Create one additional chart that answers a public health question. State the question, justify the chart, use an informative title, and add alt text.

```{r}
#| label: independent-chart
gpm_07id <- cont_sum %>%
  ggplot(
    aes(x = continent, 
        y = mean_life_expectancy,
        )) +
  geom_col(fill = "pink") +
  labs(
    title = "Average Life Expectancy by Continent in 2007",
    x = "Continent",
    y = "Average Life Expectancy (years)",
    alt = "Bar chart comparing average life expectancy across continents in 2007. Europe and Oceania have the highest average life expectancy, while Africa has the lowest."
  ) +
  theme_bw()
gpm_07id
```

## Independent Visualization Information 
How does average life expectancy differ across continents in 2007?
A bar chart is appropriate because the question compares average life expectancy across the continents groups. The bar chart can provide one simple finding without complex comparisons. 

# Reflection

1. Which aesthetic mappings changed the meaning of the graph?
Once manipulating the independent and dependent variables (x and y), this can change the interpretation of the graphs findings. For example, if the GDP per Capita (x) and the Life Expectancy (y), the graph will showcase how Life Expectancy is affected by GDP per Capita. If you were to flip these, the graph will highlight how GDP per capita influences the Life Expectancy.

2. When did faceting improve interpretation?
Faceting improved interpretation by separating the continents and providing a space clear demonstration on the patterns within each continent without overlap. 

3. Which graph most clearly communicated one finding?
The bar chart with the Life Expectancy by Continent. The differences between continents were readable without needing to interpret multiple variables.
