---
title: "Lab 4: Visualization Redesign and Accessibility"
author: "Kimaya Lacey"
format: docx
editor: visual
---

# Learning Objectives

You will critique, redesign, and test a visualization for accessibility.

# Setup

Load `tidyverse` and `gapminder`.

```{r}
#| label: loading tidyverse and gapminder

library(tidyverse)
library(gapminder)

```


# Part 1: Poor Graph

Using 2007 Gapminder data, create a graph with at least five intentional weaknesses.

```{r}
#| label: poor-graph
gapminder_2007 <- gapminder %>%
  filter(year == 2007) 

ggplot(gapminder_2007, 
       aes(
         x = gdpPercap, 
         y = lifeExp, color = continent
         )) +
  geom_point(size = 5) +
  geom_hline(yintercept = 80, linetype = "dashed") +
  labs(
    title = "Countries in 2007",
    x = "GDP",
    y = "Life"
  )
```

# Part 2: Critique

Identify five problems and explain their effects.

1. The title "Countries in 2007" does not explain to the reader what variables are being compared or the ideation behind the graph. 
2. The x-axis label only says "GDP", which is unclear for the reader because it represents GDP per capita and can be mistaken for another label. 
3. The y-axis label only says "life" and the reader would not be able to understand that the variable represents life expectancy. 
4. The points are huge and overlapping each other and this can make reading the graph difficult and readers can't distinguish data between the continents. 
5. The y-axis begins at 40 rather than 0 and this scale can make differences in life expectancy appear larger than they actually are.


# Part 3: Redesign

Create an improved graph with a finding-focused title, meaningful order, accessible design, clear labels, reduced clutter, alt text, and source.

```{r}
#| label: redesigned-graph

ggplot(
  gapminder_2007,
  aes(
    x = gdpPercap, 
    y = lifeExp, colour = continent)
) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_log10() +
  labs(
    title = "GDP per Capita Associated With Life Expectancy in 2007",
    subtitle = "Source: Gapminder, 2007",
    x = "GDP per capita",
    y = "Life expectancy (years)",
  ) +
  theme_classic()
```

# Part 4: Highlighting

Create a graph that highlights one continent and mutes the others.

```{r}
#| label: highlight-version

ggplot(
  gapminder_2007,
  aes(
    x = gdpPercap, 
    y = lifeExp, fill = continent)
) +
  geom_point(
    data = gapminder_2007 %>% 
      filter(continent != "Africa"),
    size = 2,
    alpha = 0.25,
  ) +
  geom_point(
    data = gapminder_2007 %>% 
      filter(continent == "Africa"),
    size = 3,
    alpha = 1, 
  ) +
  scale_x_log10() +
  labs(
    title = "Africa Has Lower Life Expectancy compared to other continents in 2007",
    x = "GDP per capita",
    y = "Life expectancy (years)",
    subtitle = "Source: Gapminder, 2007"
  ) +
  theme_classic()
```

# Part 5: Accessibility Review

Evaluate color dependence, contrast, font size, legend use, grayscale readability, and alt text.

**Response:**  

# Part 6: Published Graphic Critique

Critique an instructor-approved public health graphic for purpose, data, chart choice, accessibility, uncertainty, and ethics.

**Response:**
