---
title: "Lab 3: Data Wrangling for Visualization"
subtitle: "Student Version"
format:
  html:
    toc: true
    toc-depth: 3
    code-copy: true
    code-tools: true
execute:
  eval: false
  warning: false
  message: false
---

# Learning Objectives

You will clean, reshape, summarize, and join public health data before visualization.

# Setup

Load `tidyverse`.

```{r}
#| label: load-packages
library(tidyverse)
```

# Part 1: Create Raw Data

Create `raw_health_data` with at least eight counties, region, 2023 and 2024 population and case columns, one duplicate county, and one missing value.

```{r}
#| label: raw-data
raw_health_data <- data.frame(
  county = c(
    "Fulton", "DeKalb", "Cobb", "Gwinnett", "Clayton",
    "Chatham", "Richmond", "Muscogee", "Bibb", "Fulton"
  ),
  
  region = c(
    "Metro Atlanta", "Metro Atlanta", "Metro Atlanta", "Metro Atlanta",
    "Metro Atlanta", "Coastal", "Central Savannah River",
    "West Central", "Central", "Metro Atlanta"
  ),
  
  population_2023 = c(
    1091240, 764400, 787100, 971300, 306800,
    295100, 207400, 202600, 157800, 1091240
  ),
  
  population_2024 = c(
    1103500, 772100, 794800, 983600, 309200,
    298400, 209100, 204300, 159100, 1103500
  ),
  
  cases_2023 = c(
    842, 615, 487, 721, 268,
    193, 156, 141, 119, 842
  ),
  
  cases_2024 = c(
    901, 643, 512, 756, 291,
    207, 164, 153, NA, 901
  )
)

raw_health_data
```

# Part 2: Diagnose

Use `glimpse()`, duplicate checks, and missing-value summaries.

```{r}
#| label: diagnose-data

glimpse(raw_health_data)
```

# Part 3: Remove Duplicates

Create `health_data_clean` with one row per county.

```{r}
#| label: remove-duplicates

health_data_clean <- raw_health_data %>%
  distinct(county, .keep_all = TRUE)
health_data_clean
```

# Part 4: Reshape Long

Create county, region, year, population, and cases columns.

```{r}
#| label: reshape-long

health_data_long <- health_data_clean %>%
  pivot_longer(
    cols = c(population_2023, population_2024,
             cases_2023, cases_2024),
    names_to = c(".value", "year"),
    names_pattern = "(population|cases)_(\\d{4})"
  )

health_data_long
```

# Part 5: Calculate Rates

Calculate cases per 100,000.

```{r}
#| label: calculate-rates
health_data_long <- health_data_long %>%
  mutate(
    cases_per_100k = cases / 100000
  )
# Add your code here.
```

# Part 6: Summarize

Calculate mean rate by region and year.

```{r}
#| label: summarize-region

health_data_sum <- health_data_long %>%
  mutate(
    rate = (cases / population) * 100000
  ) %>%
  group_by(region, year) %>%
  summarize(
    mean_rate = mean(rate, na.rm = TRUE),
    .groups = "drop"
  )
health_data_sum
# Add your code here.
```

# Part 7: Join Contextual Data

Create a county rural-status table and join it with `left_join()`. Check duplicates, unmatched rows, and row counts.

```{r}
#| label: join-data
county_rural_status <- data.frame(
  county = c(
     "Fulton", "DeKalb", "Cobb", "Gwinnett", "Clayton",
    "Chatham", "Richmond", "Muscogee", "Bibb"
  ),
  rural_status = c(
    "Urban", "Rural", "Rural", "Urban", "Urban",
    "Urban", "Rural", "Urban", "Rural"
  )
)

county_health_join <- health_data_clean %>%
  left_join(
    county_rural_status,
    by = "county"
  )

```

# Part 8: Visualize

Create a regional line graph, ranked 2024 county graph, and rural-status comparison.

```{r}
#| label: Regional line visulaization
regional_rates <- health_data_long %>%
  mutate(
    rate = (cases / population) * 100000
  ) %>%
  group_by(region, year) %>%
  summarize(
    mean_rate = mean(rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  regional_rates,
  aes(
    x = year,
    y = mean_rate,
    group = region,
    color = region
  )
) +
  geom_line(
    linewidth = 1) +
  geom_point(
    size = 3) +
  labs(
    title = "Mean Health Case Rate by Region",
    subtitle = "Comparing regional case rates from 2023 to 2024",
    x = "Year",
    y = "Mean Cases per 100,000",
    color = "Region"
  ) +
  theme_classic()
```


```{r}
#| label: Ranked 2024 county visualization
county_ru24 <- county_health_join %>% 
  select(county, population_2024, cases_2024) %>%
  mutate(
    rate = (cases_2024 / population_2024) * 100000
  ) %>%
  arrange(desc(rate))

ggplot(
  county_ru24,
  aes(
    x = reorder(county, rate), 
    y = rate
  )
) +
  geom_col(fill = "pink") +
  coord_flip() +
  labs(
    title = "2024 Health Case Rate by County",
    subtitle = "County-level comparison of reported cases per 100,000 residents in 2024",
    x = "County",
    y = "Cases per 100,000"
  ) +
  theme_classic()
```


```{r}
#| label: Rural Status Comparison Visualization

rural_compar <- county_health_join %>%
  drop_na(cases_2023, population_2023, rural_status) %>%
  mutate(
    rate = (cases_2023 / population_2023) * 100000
  ) %>%
  group_by(rural_status) %>%
  summarize(
    mean_rate = mean(rate, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(
  rural_compar,
  aes(
    x = rural_status,
    y = mean_rate, fill = rural_status
  )
) +
  geom_col() +
  labs(
    title = "2023 Health Case Rate by Rural Status",
    subtitle = "Comparing mean case rates between rural and urban counties",
    x = "Rural Status",
    y = "Mean Cases per 100,000"
  ) +
  theme_classic()
```

# Reflection

Explain why long format was useful, what causes row multiplication during joins, and why rates are preferable to counts.

