---
title: "Lab 1: Reproducible Data Visualization Workflow"
author: "Kimaya Lacey"
format: docx
editor: visual
---

## Part 3: Loading Packages

```{r}
#| label: load-packages
install.packages("tidyverse")
install.packages("here")
install.packages("fs")

library(tidyverse)
library(here)
library(fs)
```

## Part 4: Confrim the project Root

```{r}
#| label: confirm-project-root
here()

```

# Answer:

1.  What path was printed?
2.  Does it point to the correct course project?
3.  Why is working from an R Project preferable to manually setting the working directory?

## Part 5: Create the Folder Structure

```{r}
#| label: create-folder-structure

folders <- c(
  "data",
  "figure",
  "scripts",
  "tables",
  "assignments",
  "labs",
  "portfolio"
)

for (folder in folders) {
  dir.create(here::here(folder), showWarnings = FALSE)
}
```

# Part 6: Inspect the Folder Structure

```{r}
#| label: inspect-folder-structure
dir_ls()
```

# Part 7: Create a Simulated Public Health Dataset

Create a reproducible county-level dataset with the following variables:

-   `county`
-   `region`
-   `adult_population`
-   `asthma_visits`
-   `vaccination_rate`
-   `asthma_visit_rate`

Requirements:

-   Include 12 fictional counties.
-   Include three regions: North, Central, and South.
-   Use `set.seed(716)`.
-   Generate plausible values for population, visits, and vaccination coverage.
-   Calculate asthma visit rate per 100,000.
-   Round the rate to one decimal place.
-   Use tidy formatting.

```{r}
#| label: create-simulated-health-data
publichealth_county <- data.frame(
  county = c("Cook", "Lee", "Dekalb", "Lithonia", "Rockdale", "Cobb", "Fulton", "Gwinett", "Chatham", "Hill", "Richmond", "Henry"), 
  region = c("North", "Central", "South", "South", "North", "Central", "Central", "South", "North", "South", "Central", "North"),
  adult_population = c("245,000", "746,218", "501,321", "175,093", "632,844", "382,742", "946,129", "123,456", "789,032", "216,822", "876,543", "497,641"),
  vaccination_rate = c(".92",".14",".67",".53",".81",".30",".49",".25",".78", ".42",".56",".29" ),
  asthma_visit_rate = c("245.6", "318.4", "187.9","421.7","356.2","279.5","512.3", "198.7","384.1","267.8","449.6","331.5")
)

publichealth_county
```

# Part 8: Save the Dataset

Save the dataset as:

``` text
data/county_health_data.csv
```

Use `write_csv()` and `here()`.

```{r}
#| label: save-simulated-health-data
#| data/publichealth_county.csv
write_csv(publichealth_county, "publichealth_county.csv")
here()
```

# Part 9: Remove and Re-Import the Dataset

Remove the object from the environment.

```{r}
#| label: remove-health-data
rm(publichealth_county)
```

Then re-import the CSV file using `read_csv()` and `here()`.

```{r}
#| label: import-health-data
publichealth_county <- read_csv(
  here("publichealth_county.csv")
)
```

# Part 10: Inspect the Data

Use `glimpse()`.

```{r}
#| label: inspect-health-data
glimpse(publichealth_county)
```

Create a one-row summary containing:

-   number of counties,
-   mean vaccination rate,
-   median asthma visit rate.

```{r}
#| label: summarize-health-data

phc_sum <- publichealth_county %>%
  summarise(
    num_county = n_distinct(county),
    mean_vaccination_rate = mean(vaccination_rate),
    median_asthma_visit_rate = median(asthma_visit_rate)
  )

phc_sum
```

Answer:

1.  What does one row represent? Descriptive Statistics
2.  How many variables are included? 3
3.  Which variables are categorical? Region
4.  Which variables are numeric? Vaccination rate, population, and asthma visit
5.  Which variable should be used to compare asthma burden fairly across counties? Region

# Part 11: Create a Ranked Vaccination Bar Chart

Create a horizontal bar chart ranking counties by vaccination coverage.

Requirements:

-   Reorder counties by vaccination rate.
-   Add a dashed reference line at 80%.
-   Use an informative title and subtitle.
-   Label the y-axis as vaccination coverage.
-   State that the values are simulated.
-   Use a readable theme.
-   Store the graph as `vaccination_plot`.

```{r}
#| label: create-vaccination-plot
#| fig-cap: "Simulated vaccination coverage by county."
#| fig-alt: "Horizontal bar chart ranking fictional counties by simulated vaccination coverage."
vaccination_plot <- publichealth_county %>%
                    ggplot(
                      aes(
                      x = county,
                      y = vaccination_rate, colour = region
                    )) + 
  geom_col(fill = "pink") + 
  geom_hline(
    yintercept = .80,
    linetype = "dashed"
    ) + 
  coord_flip() +
labs(
  title = "County Vaccination Coverage", 
  subtitle = "Simulated vaccination coverage by county", 
  x = "County",
  y = "Vaccination Rate"
) + 
  theme_minimal()
vaccination_plot
```

# Part 12: Interpret the Figure

Answer in two or three complete sentences.

1.  What pattern does the figure show?
2.  Which counties appear furthest below the target?
3.  What context would be required before using this graph for a real decision?

**Response:**

# Part 13: Save the Vaccination Figure

Save the graph as:

``` text
figures/lab01_vaccination_coverage.png
```

Use:

-   width of 8 inches,
-   height of 6 inches,
-   resolution of 300 dpi.

```{r}
#| label: save-vaccination-plot
ggsave(
  here("figure", "lab01_vaccination_coverage.png"),
  plot = vaccination_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

# Part 14: Independent Visualization

Create one additional visualization using the same dataset.

Choose one:

-   scatterplot of adult population and asthma visits,
-   bar chart of mean vaccination coverage by region,
-   dot plot of asthma visit rates by county,
-   or another instructor-approved graph.

Requirements:

-   Store the graph as `independent_plot`.
-   Include an informative title and subtitle.
-   Label both axes clearly.
-   Add a caption identifying the data as simulated.
-   Use a readable theme.

```{r}
#| label: independent-visualization
#| fig-cap: "Independent visualization created by the student."
independent_plot <- publichealth_county %>%
                    ggplot(
                      aes(
                        x = adult_population,
                        y = asthma_visit_rate, colour = region
                      )) + 
  geom_line() + 
  labs(
    title = "Frequency of Asthma Visits by Adult Population",
    subtitle = "Independent visualization created by the student",
    x = "Adult population",
    y = "Asthma rates per 100,000 people"
  )
independent_plot
# Add your code here.
```

# Part 15: Save the Independent Figure

Save the graph as:

``` text
figures/lab01_independent_visualization.png
```

Use:

-   width of 8 inches,
-   height of 5 inches,
-   resolution of 300 dpi.

```{r}
#| label: save-independent-visualization
ggsave(
  here("figures", "lab01_independent_visualization.png"),
  plot = vaccination_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

# Part 16: Tidy Coding Check

Review your code and confirm:

-   [ ] `%>%` is used instead of the native pipe
-   [ ] one major function appears per line
-   [ ] arguments are indented consistently
-   [ ] object names are descriptive
-   [ ] no `setwd()` appears
-   [ ] no absolute file paths appear
-   [ ] code chunks have labels

# Part 17: Git and GitHub

Complete the following:

1.  Open the Git pane.
2.  Review the changed files.
3.  Stage the files associated with this lab.
4.  Commit using a meaningful message.
5.  Push to GitHub.
6.  Open the repository online and verify that the files appear.

Suggested commit message:

``` text
Complete Lab 1 reproducible workflow
```

# Reflection Questions

Answer each question in complete sentences.

## Question 1

Why is reproducibility important in public health research and practice?

**Response:**

## Question 2

How does a consistent project structure improve collaboration and review?

**Response:**

## Question 3

Why are relative paths preferable to absolute paths?

**Response:**

## Question 4

How does GitHub contribute to reproducibility?

**Response:**

## Question 5

What was the most challenging part of the lab?

**Response:**

# Deliverables
