---
title: "Lab 1: Reproducible Data Visualization Workflow"
author: "Kimaya Lacey"
format: docx
editor: visual
---

## Part 3: Loading Packages

```{r}
#| label: load-packages
#| include: false
#install.packages("tidyverse")
#install.packages("here")
#install.packages("fs")

library(tidyverse)
library(here)
library(fs)
```

## Part 4: Confrim the project Root

```{r}
#| label: confirm-project-root
#| echo: false
here()
```

# Answer:

1.  What path was printed?" /Users/klacey/Downloads/Data Visualizations"
2.  Does it point to the correct course project? Yes
3.  Why is working from an R Project preferable to manually setting the working directory? The R Project will provide easy access to all documents, figures, and data essential for the project.

## Part 5: Create the Folder Structure

```{r}
#| label: create-folder-structure
#| echo: false

folders <- c(
  "data",
  "figure",
  "scripts",
  "tables",
  "assignments",
  "labs",
  "portfolio"
)

for (folder in folders) {
  dir.create(here::here(folder), showWarnings = FALSE)
}
```

# Part 6: Inspect the Folder Structure

```{r}
#| label: inspect-folder-structure
#| echo: false
dir_ls()
```

# Part 7: Create a Simulated Public Health Dataset

```{r}
#| label: create-simulated-health-data
#| echo: false
publichealth_county <- data.frame(
  county = c("Cook", "Lee", "Dekalb", "Lithonia", "Rockdale", "Cobb", "Fulton", "Gwinett", "Chatham", "Hill", "Richmond", "Henry"), 
  region = c("North", "Central", "South", "South", "North", "Central", "Central", "South", "North", "South", "Central", "North"),
  adult_population = c("245,000", "746,218", "501,321", "175,093", "632,844", "382,742", "946,129", "123,456", "789,032", "216,822", "876,543", "497,641"),
  vaccination_rate = c(".92",".14",".67",".53",".81",".30",".49",".25",".78", ".42",".56",".29" ),
  asthma_visit_rate = c("245.6", "318.4", "187.9","421.7","356.2","279.5","512.3", "198.7","384.1","267.8","449.6","331.5")
)

publichealth_county
```

# Part 8: Save the Dataset

```{r}
#| label: save-simulated-health-data
write_csv(publichealth_county, "publichealth_county.csv")
here()
```

# Part 9: Remove and Re-Import the Dataset

```{r}
#| label: remove-health-data
#| echo: false
rm(publichealth_county)
```

```{r}
#| label: import-health-data
#| echo: false
publichealth_county <- read_csv(
  here("publichealth_county.csv")
)
```

# Part 10: Inspect the Data

```{r}
#| label: inspect-health-data
#| echo: false
glimpse(publichealth_county)
```

```{r}
#| label: summarize-health-data
#| echo: false

phc_sum <- publichealth_county %>%
  summarise(
    num_county = n_distinct(county),
    mean_vaccination_rate = mean(vaccination_rate),
    median_asthma_visit_rate = median(asthma_visit_rate)
  )

phc_sum
```

Answer:

1.  What does one row represent? One County and its corresponding metrics.
2.  How many variables are included? 3
3.  Which variables are categorical? Region and County
4.  Which variables are numeric? Vaccination rate, population, and asthma visit
5.  Which variable should be used to compare asthma burden fairly across counties? Asthma Visit Rate as it is reported per 100,000 people across each county.

# Part 11: Create a Ranked Vaccination Bar Chart

```{r}
#| label: create-vaccination-plot
#| echo: false
#| fig-alt: Horizontal bar chart ranking fictional counties by simulated vaccination coverage.
#| fig-cap: Simulated vaccination coverage by county.
vaccination_plot <- publichealth_county %>%
                    ggplot(
                      aes(
                      x = county,
                      y = vaccination_rate, colour = region
                    )) + 
  geom_col(fill = "pink") + 
  geom_hline(
    yintercept = .80,
    linetype = "dashed"
    ) + 
  coord_flip() +
labs(
  title = "County Vaccination Coverage", 
  subtitle = "Simulated vaccination coverage by county", 
  x = "County",
  y = "Vaccination Rate"
) + 
  theme_minimal()
vaccination_plot
```

# Part 12: Interpret the Figure

Answer in two or three complete sentences.

1.  What pattern does the figure show? The figure shows that vaccination rates vary considerably across counties. Cook and Rockdale have the highest vaccination rates and exceed the 80% target, while most other counties fall below the target.

2.  Which counties appear furthest below the target? The Lee, Cobb, Henry, and Gwinnett counties appear the furthest below the 80% line, with Lee having the lowest vaccination rate.

3.  What context would be required before using this graph for a real decision? Before using this graph for a real decision, we would need to know whether the vaccination data are reliable and current, how the rates were calculated, and which population was included. The graph is based on simulated data, real and verified data would be necessary before making public health decisions.

# Part 13: Save the Vaccination Figure

```{r}
#| label: save-vaccination-plot
#| echo: false
 ggsave(
  filename = "../figure/lab01_vaccination_coverage.png",
  plot = vaccination_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

# Part 14: Independent Visualization

```{r}
#| label: independent-visualization
#| echo: false
#| fig-cap: Independent visualization created by the student.
independent_plot <-publichealth_county %>%
      ggplot(
        aes(
        x = adult_population,
        y = asthma_visit_rate,
        colour = region
    )
  ) +
  geom_point() +
  scale_color_viridis_d() +
  labs(
    title = "Frequency of Asthma Visits by Adult Population",
    subtitle = "Simulated Asthma rates by adult population",
    x = "Adult population",
    y = "Asthma visits per 100,000 people",
    colour = "Region"
  ) +
  theme_minimal()

independent_plot
# Add your code here.
```

# Part 15: Save the Independent Figure

```{r}
#| label: save-independent-visualization
#| echo: false
 ggsave(
  filename = "../figure/lab01_independent_visualization.png",
  plot = vaccination_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

# Reflection Questions

## Question 1

Why is reproducibility important in public health research and practice?

Reproducibility is important in public health research and practice because when working with data, trust and credibility is critical. Public Health practitioners should be able to access current projects/data with no differences from the original file. Therefore allowing practitioners to analyze the same document for either similar or different subjects.

## Question 2

How does a consistent project structure improve collaboration and review?

A consistent project structure improves collaboration and review by providing an simple method in communicating research findings to peers and non-scientific audience (if necessary). Consistency is key when working with tedious project platforms, such as RStudio foe example.

## Question 3

Why are relative paths preferable to absolute paths?

Relative paths are preferable because they reference files directly from the project folder rather than a specific computer's file location/path. This makes the project easier to share, reproduce, and run on another device without changing the file directory.

## Question 4

How does GitHub contribute to reproducibility?

GitHub contributes to reproducibility by storing the project's code, data, and documentation on one site. Additionally, it provides a history of edits, allowing different users to review previous versions and recreate the analysis using the same materials.

## Question 5

What was the most challenging part of the lab?

The most challenging part of the lab was pushing my R project file to Github. Of course I needed a refresher from previous lessons from last semester, but this assignment helped rekindle those useful skills!

# Stretch Challenge

Improve one graph by adding one of the following:

-   a colorblind-friendly palette

Write two sentences explaining why the change improves communication.

The colorblind-friendly palette improves accessibility by including color-blind friendly colors between regions easier to distinguish for viewers with color-blindness. This change improves communication by expanding the audience to thoroughly interpret the regional patterns in asthma visit rates.
